How unprecedented was the February 2021 Texas cold snap?

### 2021-04-23 doc/review_response.tex

\documentclass{ar2rc}

\usepackage{url}
\usepackage{siunitx}

\usepackage[colorinlistoftodos]{todonotes}
\usepackage{color}
\newcommand{\jdg}[1]{\todo[color=red!30]{#1}}
\newcommand{\all}[1]{\todo[color=blue!30]{#1}}

\hypersetup{hidelinks}

\title{How unprecedented was the February 2021 Texas cold snap?}
\author{James Doss-Gollin, David J. Farnham, Upmanu Lall, and Vijay Modi}
\journal{Environmental Research Letters}

\begin{document}

\maketitle

\listoftodos

\section{Referee \#1}

\RC  This is an excellent, timely analysis. The authors are right to point to the need for more in depth studies on how cascading failures occurred during the 2021 Texas freeze, and the steps required to minimize their impacts in the future. But this is a much needed, early contribution, and very well done. I don't have any major criticisms of the work, which is well organized and rigorous.

\AR Thank you for your comments. We agree that this points to a need for more in depth studies, particularly related to cascading failures of natural gas and electricity systems but also to failure of water and other systems.

\subsection{Some minor comments:}

\RC Considering I think this paper will be widely read and cited, one general comment is that I felt like the citations were a little light in the introduction, especially regarding the 2021 freeze itself.

\AR This is a point that reviewer \#2 has also brought up.\all{add some citations}

\RC `Texas interconnect' (which I assume is the Texas Interconnection bulk electric power system administered by ERCOT) is used several times but (I don't think) ever defined; many folks will not know what this refers to. If you added a brief definition it would help. If this does refer to the grid, then an additional comment is that I've never heard it referred to as the Interconnect, but rather the Interconnection.

\AR This is a helpful comment. The Interconnection is a better term and we have revised it throughout.\jdg{Change language to always say TX Interconnection}\jdg{define Texas Interconnection}. This will clarify the document and facilitate understanding.

\RC Regarding the definition of a HDD/CDD: More frequently 65 degrees is used, I think.

\AR Do we need to do any analysis, or just reply that it doesn't matter?\all{Assess}

\RC On the difficulty of estimating supply side risk -- is it possible at this point to map the locations of plants that lost functionality?

\AR This is a helpful suggestion. When we submitted the manuscript, this data was not available. However, ERCOT has since released data on generation failure.\jdg{Download this data and update the map}

\RC In the discussion, you mention the 1983 event as being one of comparable severity, and then talk about the grid's failure in 2021. I don't know if this was your intention, but  when I read this sentence the first time I took it to mean the grid is *more* vulnerable now. I think you could reasonably make that case with greater electrification of heating. But I also wondered about coal-to-gas switching, and whether 1983 would have likely been a different fleet of power plants that were perhaps less likely to lose useable capacity during cold weather.

\AR Thank you for the interesting and thought-provoking comment.\all{Discuss: do we have thoughts?}

\section{Referee \#2}

\RC This paper investigates whether the Texas deep freeze event was a black swan event, or if it should have been predicted from previous cold snaps. The authors compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for heating demand and use that to determine aggregate heating demand for 2021. Overall the authors find that demand for previous storms was more intense than the 2021 disaster.

\AR Thank you for this helpful summary.

\RC The introduction is seriously lacking in citations and references. For example, this line ``While polar air excursions are not unusual, they do not typically reach as far south as the Mexican border.'' And a number of others need citations and would benefit from concrete numbers. A question that arises is how many storms have reached the Mexican border in the last 10 years? Line 45-46 on page 1 also needs a citation.

\AR Thank you for this comment. We note that referee \#1 made a similar comment. We have now added refererences\ldots\all{Add more references as noted above}

\RC In the introduction the ``population weighted temperature excursion below \SI{68}{\degree F}'' needs a definition for temperature excursion. Figure 1 is missing citations for the data. What do you mean by days ending? Were these temperatures recorded at a specific time of day? Figure 1 would be improved by adding column headers for the averages (1-day, 3 day, and 5 day).

\AR Thank you for calling to our attention these unclear points. First, we have changed the subplot titles of Figure 1 from ``N days ending YYYY-MM-DD'' to ``YYYY-MM-DD to YYYY-MM-DD'' for clarity. We have supplemented this with column titles. We have also added a reference to the ERA-5 data in the caption.\jdg{update the figure as described}

\RC A weakness of this paper is the lack of comment about the capacity and size of the Texas grid. How does the inferred demand compare to the size of the ERCOT power grid and the interconnection capability over time? We know that the ERCOT grid is larger in 2021 than in 1989, but from Figure 2 there should have been more demand in 1989 than in 2021. Would most of the people have wood stoves? Was the blackout and number of people impacted as large in the past, as it was in 2021?

\AR Thank you for bringing this point to our attention. To minimize confusion, we have clarified our terminology so that ``inferred heating demand'' is not ``inferred heating demand per capita.'' This is reflected throughout the document.\jdg{Update language throughout} However, comparing the actual infrastructure failures across different events (such as 1989) would require understanding and modeling cascading failures of interdependent infrastructure systems, which is beyond the scope of our analysis.

\RC On page 7 line 51 the authors say the Oklahoma recorded some water main breaks and rolling blackouts, but infrastructure impacts were less severe. To strengthen this paper, the authors should provide insights as to why this was the case for OK, and not TX.

\AR Perhaps ``it gets cold there'' but I'd hate to say things without any evidence?\all{Comment}

\RC I understand that a big takeaway from the paper is that the cold temperatures and heat demand could be predicted since there were similar demand numbers from 1989, but one year of similar demand does not eliminate this as a black swan event. From the discussion period it seems that over a 30-year time period this type of event has only happened twice. That seems very rare in my opinion.

\AR Thank you for this comment. Whether it was a ``black swan'' or not depends on a definition of black swan. We have provided a definition and citation for this term.\jdg{Cite Taleb and add definition} Beyond this one instance, we have sought to clarify our language to ask only whether the temperatures were ``unprecedented.'' Since we find that, particularly at the aggregate scale, there were historical events of comparable intensity and duration, we can state that the event ``was not unprecedented.'' We also attempt to quantify return periods. However, we have removed references to terms like ``rare'' or ``surprising'' \jdg{check for these terms} that do not have specific interpretations to minimize any confusion.

\RC The conclusions says that several other storms would have resulted in similar demand. How many?

\AR Thank you for pushing us to be more specific with our claims. The text now reads (\ldots).

\RC This paper needs to tie in the size and capacity of the Texas grid to the demand projections. Without it I believe the claims are overstated.

\AR Thank you for this comment. The first step we have taken is, as noted above, to refer to ``inferred demand for heating per capita'' to emphasize that we are not accounting for population change.\all{anything else we should include?}

\end{document}

### 2021-04-23 doc/submission_02.tex
address some reviewer comments -- discussion still needed for others

\vspace*{1in}

{\LARGE How unprecedented was the February 2021 Texas cold snap?}

\vspace*{1in}

This is a non-peer reviewed manuscript by James Doss-Gollin, David J. Farnham, Upmanu Lall, and Vijay Modi.
You can find a permanent and updated preprint on EarthArXiV at \url{https://doi.org/10.31223/X5003J}.
This manuscript is still under review; if accepted we will provide a link to the published article here.

\clearpage

%\title[PREPRINT: Doss-Gollin, Farnham, Lall, and Modi]{How unprecedented was the February 2021 Texas cold snap?}
\title[PREPRINT by Doss-Gollin, Farnham, Modi, and Lall]{How unprecedented was the February 2021 Texas cold snap?}

@@

\begin{abstract}
  Winter storm Uri brought severe cold to the southern United States in February 2021, causing a cascading failure of interdependent systems in Texas where infrastructure was not adequately prepared for such cold.
  In particular, the failure of interconnected energy systems reduced electricity supply just as heating demands spiked, leaving millions of Texans without heat or electricity, many for several days.
  This motivates the question: was the cold that contributed to this infrastructure failure a ``black swan'' that could not have been anticipated, or did historical storms provide a precedent?
  We compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for heating demand and use this metric to answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred today?''.
  We find that local temperatures and the inferred demand for heating across the Texas Interconnect during a storm in December 1989 were more intense than those recorded during February 2021, and that several other storms in the modern era were comparable.
  Given anticipated population growth, future storms may lead to even greater infrastructure failures if adaptive investments are not made.
  Further, electricity system managers should anticipate that upward trends in electrification of heating may cause peak annual loads on the Texas Interconnect to occur during winter storms.

@@

Since production and distribution of natural gas and electricity is possible under conditions far colder than any Texas experienced in February 2021, energy system failures reflect inadequate preparedness for cold.
If this were an unprecedented or very rare event then it is possible that the impacts resulted from an event outside the design envelope.
On the other hand, if similar events were in the record then one would expect a higher level of preparation.
It is therefore important to assess whether the February 2021 cold snap could have been anticipated.
To answer this question, we first compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for the unknown heating demand, and use this metric to explore how rare this event was using historical data.
Next, we conduct a spatially distributed analysis to assess how unexpected the cold experienced by local roads, water mains, gas pipelines, energy generation facilities, and critical infrastructure installations was across Texas.
We conclude by discussing the implications of these findings for long-term electricity systems planning given anticipated population growth and electrification.

@@

We use three distinct datasets to analyze temperature minima in Texas through the lens of distributed (each grid cell analyzed separately) and aggregated (weighted averages taken across space) extreme values analysis.

@@

We use three temperature datasets to ensure robust findings:
\begin{enumerate}
  \item Hourly \SI{2}{\meter} air temperature reanalysis on a \SI{0.25}{\degree} grid from the ERA-5 reanalysis project produced by the European Centre for Medium Range Weather Forecasting \cite{hersbach_era5:2020} and available  from the Copernicus Data Store (\url{https://cds.climate.copernicus.eu}) from 1950 to the present.
        The period from 1950 to 1979 is released as a preliminary back extension.
        All plots shown in the main text use the ERA-5 data, but supplemental figures use other data sets.
  \item Daily mean, minimum and maximum temperatures, gridded to \SI{1}{\degree}, produced by Berkeley Earth (\url{http://berkeleyearth.org/data/}).
        This gridded product is based on statistical analysis of station data and is available from 1880 to 2019.
        This dataset is considered an experimental product, so we use it only for comparative purposes.
  \item To complement blended gridded data products, we use station temperature data from the Global Historical Climatology Network (GHCN) dataset compiled by the National Ocean and Atmospheric Administration \cite{Menne:2012hk} and available at \url{https://www1.ncdc.noaa.gov/pub/data/ghcn/daily/}.
        This dataset provides daily mean, maximum, and minimum temperature observations.
        These measurements represent point measurements, which can differ in important ways from gridded products describing spatial averages due to the spatial heterogeneity of temperature fields.
        We retain stations within the state of Texas if they provide at least 60 years of data and if they contain observations for the set of historical cold extremes shown in \cref{fig:historic_era5}.
\end{enumerate}
We also use population density data from the GPWv4 dataset \cite{ciesin_gpwv4:2016}, a list of power generation facilities from the US Energy Information Administration \cite{useia_generators:2021}, and a map of the Texas Interconnect \cite{useia_regions:2021}.

\subsection{Inferred Demand for Heating}\label{sec:inferred-demand}

Most space heating in Texas is either electric or gas \cite{waite_heating:2020} and the majority of power generation in the Texas Interconnect depends on natural gas \cite{everhart_iea:2021}.
Stress on natural gas production and delivery was therefore just as important as the more visible stress on the electric system.

@@

To assess how spatially correlated cold spells might affect the Texas electric grid, we average heating demand in space over the Texas Interconnect domain \cite{useia_regions:2021}, weighting each grid cell by 2020 population density \cite{ciesin_gpwv4:2016}.
We refer to this spatially aggregated time series, which has the straightforward interpretation as the average heating demand experienced by Texas residents, as ``Inferred Demand for Heating.''

@@

For each event duration considered, return periods are obtained by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual minimum December-February (DJF) temperatures, or to the time series of annual maxima of inferred demand for heating (\cref{sec:inferred-demand}).
Events that occur in December are coded to the following year.
The 2021 winter season was excluded from return period estimates, allowing us to interpret return periods for the February 2021 event as \emph{a priori} estimates.

@@

\section{How extreme was inferred heating demand over the Texas Interconnect?}

The total shock to Texas heating demand is partially determined by the extent to which cold snaps impact multiple population centers simultaneously.
As such, understanding whether there was precedent for a cold snap simultaneously affecting several regions of Texas's grid that today have high population density is critical.
We therefore use our measure of inferred demand for heating (see \cref{sec:inferred-demand}) to represent the aggregate heating demand induced by cold temperatures.
Aggregating historic temperature fields in space using the 2020 population, we answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred today?''

\Cref{fig:idf_weighted} shows that the intensity, duration, and recurrence intervals of the February 2021 storm are severe but not unprecedented in the historical record.
For example, at the 6 hour duration the December 1989 storm was substantially more intense and other storms including February 1951 were nearly as intense.
At the two day duration, the 2021 and 1989 events were approximately equally intense and other storms including December 1983 were nearly as intense.
The 2011 storm, which caused rolling blackouts and motivated research into the energy system's vulnerability to cold \cite{ferc_outages:2011}, was quite modest by comparison.
The right panel shows statistical return periods for these extreme events, though these estimates are sensitive to small changes in inferred demand for heating.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_MLE_popweighted.pdf}
  \caption{
    The inferred demand for heating induced by the February 2021 cold snap is not unprecedented.
    For the worst 6 hours, the 1989 event was more severe than the 2021 event, while they are comparable for the longer durations.
    (a): time series of annual maximum inferred demand for heating (\cref{sec:inferred-demand}) at 6 hour and 2 day durations.
    December extremes, including the December 1989 storm, are coded to the following year so that one maximum per December-February winter season is taken.
    (b): the intensity-duration-frequency intervals estimated using 1950-2020 data (i.e., not using the 2021 event), overlaid by the annual maxima from the 1989, 2011, and 2021 events.
    Gray dashed lines indicate 2, 5, 10, 25, 50, and 100 year return levels.
  }\label{fig:idf_weighted}
\end{figure}

@@

\Cref{fig:local_era5} shows local return periods for the February 2021 cold snap at 6 hour, 1 day, 2 day, and 4 day durations.
Other than a band from south-central to south-east Texas, nearly all regions of the Texas Interconnect (gray outline in \cref{fig:local_era5}a,d) experienced cold with a return period below 50 years.
Results are similar using station data (\cref{fig:local_ghcnd}).
Importantly for the energy system, the band experiencing cold with return period greater than 50 years includes a substantial fraction of Texas’s population (\cref{fig:local_era5}a) and natural gas generation (\cref{fig:local_era5}d).
Outside the Texas Interconnect region, much of the Texas Panhandle and central Oklahoma experienced intense cold with a return period greater than 100 years at multiple durations.
Although Oklahoma recorded some water main breaks \cite{crum_water:2021} and rolling blackouts for several hours \cite{money_oklahoma:2021}, infrastructure impacts were less severe than in Texas despite recording more extreme (relative both to its own historical record and in absolute terms) temperatures.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{local_rt_era5.pdf}
  \caption{
    Return periods for the February 2021 event, calculated using stationary estimates of annual extremes over the period 1950-2020.
    Return periods are calculated separately for each cell.
    (a): estimates of 2020 population density \cite{ciesin_gpwv4:2016}.
    (d): energy generation facilities in Texas \cite{useia_generators:2021}.
    (b,c,e,f): local return periods for 6 hour, 1 day, 2 day, and 4 day durations, respectively.
    Contours enclose regions that recorded 50 and 100 year return levels.
    The gray region in panels (a) and (d) shows boundaries of the Texas Interconnect \cite{useia_regions:2021}.
  }\label{fig:local_era5}
\end{figure}

@@

Our spatially aggregated metric of inferred demand for heating shows that the February 2021 event was intense but not unforeseeable (\cref{fig:idf_weighted}).
Although specific locations experienced very intense ($>100$ year return period) temperatures, we find that for most locations in Texas the temperatures recorded during the February 2021 cold snap had precedent in the historical record and could have been anticipated given available data.

A proximate cause of load shedding in the Texas Interconnect during February 2021 was the vulnerability of the electricity generation system to the severe cold temperatures \cite{everhart_iea:2021}.
\Cref{fig:local_era5,fig:local_ghcnd} show that this happened even though most parts of the state had previously experienced similarly intense cold, notably in 1989.
While our analysis neglects other meteorological factors, like freezing rain, that may have impeded operations at specific facilities, our analysis suggests that the February 2021 failures of energy and electricity systems in the Texas Interconnect took place during temperatures comparable to those previously recorded.
Similarly, water mains broke in places like Houston where temperatures did not exceed 100-year return levels, underscoring characteristic vulnerability of critical infrastructure systems \cite{chester_reliable:2020}.

@@

Our analysis quantifies the probability with which temperature extremes could have been anticipated \emph{a priori}.
Other factors also govern infrastructure performance and failure, including precipitation, the demand for natural gas in adjacent regions, and complex connections within and between regional systems.
Similarly, decisions at multiple time scales, including disaster preparedness and risk communication, contribute to the human consequences of physical infrastructure failure.
Thus, the exact chain of events that led to the blackouts and water system disruptions during February 2021 should be sorted out only after thorough investigations by parties on the ground in Texas.

@@

The February 2021 cold snap was the most intense in 30 years, but was not without precedent in the full historical record.
In addition to the record cold conditions of 1899 (\cref{fig:historic_bk}), we estimate that the weather of December 1989 would have resulted in slightly higher 6-hour and 2-day aggregate heating demands over the Texas Interconnect had it occurred in February 2021.
Several other storms since 1950 would have produced nearly as much demand for heating.
Given upward trends in the electrification of heating, it is likely that future cold snaps will cause peak annual loads on the Texas Interconnect to occur during the winter season.
Infrastructure expansion necessitated by a rapidly growing population offers Texas the opportunity to invest in a more resilient energy system.

@@

\subsection{Inferred demand for heating}

To complement our analysis of inferred demand for heating, we consider how results change as a function of two modeling decisions.
First, we consider what happens if the spatial field demand for heating is aggregated using grid cell area rather than population density.
Next, we compute return periods using an estimator based on the method of $L$-moments.
Although $L$-moment estimators for the generalized extreme value distribution are not unbiased, they are popular in the statistical hydrology literature for their stability \cite{hosking_gev:1985,martins_gev:2001,morrison_gev:2002}.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{HDD_pop_weighted_ts.pdf}
  \caption{
    A time series of inferred demand for heating over the Texas Interconnect during the February 2021 cold snap.
  }\label{fig:hdd_ts}
\end{figure}

>>>>

\title{How unprecedented was the February 2021 Texas cold snap?}

@@

\begin{abstract}
  Winter storm Uri brought severe cold to the southern United States in February 2021, causing a cascading failure of interdependent systems in Texas where infrastructure was not adequately prepared for such cold.
  In particular, the failure of interconnected energy systems reduced electricity supply just as heating demands spiked, leaving millions of Texans without heat or electricity, many for several days.
  This motivates the question: were the temperatures that contributed to this infrastructure failure beyond what was known to be possible, or did historical storms offer a precedent?
  We compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for heating demand and use this metric to answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred today?''.
  We find that local temperatures and the inferred demand for heating across the region served by the Texas Interconnection during a storm in December 1989 were more severe than those recorded during February 2021, and that several other storms in the modern era were comparable.
  Given anticipated population growth, future storms may lead to even greater infrastructure failures if adaptive investments are not made.
  Further, electricity system managers should anticipate that upward trends in electrification of heating may cause peak annual loads on the Texas Interconnection to occur during winter storms.

@@

Since production and distribution of natural gas and electricity is possible under conditions far colder than any Texas experienced in February 2021, energy system failures reflect inadequate preparedness for cold.
If these temperatures were unprecedented in the region covered by the Texas Interconnection then it is possible that the impacts resulted from an event outside the design envelope.
On the other hand, if similar events were in the record then one would expect a higher level of preparation.
It is therefore important to assess whether historical data offered a precedent for the temperatures observed during February 2021.
To answer this question, we first compute the population weighted temperature excursion below \SI{68}{\degree F} as a proxy for the unknown heating demand, and use this metric to explore the frequency with which this event could have been expected to occur using historical data.
Next, we conduct a spatially distributed analysis to assess how unexpected the cold experienced by local roads, water mains, gas pipelines, energy generation facilities, and critical infrastructure installations was across Texas.
We conclude by discussing the implications of these findings for long-term electricity systems planning given anticipated population growth and electrification.

@@

We use three distinct datasets to analyze temperature minima in the region covered by the Texas Interconnection bulk electric power system administered by the Electric Reliability Council of Texas (hence ``Texas Interconnection'') through the lens of distributed (each grid cell analyzed separately) and aggregated (weighted averages taken across space) extreme values analysis.

@@

We use three temperature datasets to ensure robust findings:
\begin{enumerate}
  \item Hourly \SI{2}{\meter} air temperature reanalysis on a \SI{0.25}{\degree} grid from the ERA-5 reanalysis project produced by the European Centre for Medium Range Weather Forecasting \cite{hersbach_era5:2020} and available  from the Copernicus Data Store (\url{https://cds.climate.copernicus.eu}) from 1950 to the present.
        The period from 1950 to 1979 is released as a preliminary back extension.
        All plots shown in the main text use the ERA-5 data, but supplemental figures use other data sets.
  \item Daily mean, minimum and maximum temperatures, gridded to \SI{1}{\degree}, produced by Berkeley Earth (\url{http://berkeleyearth.org/data/}).
        This gridded product is based on statistical analysis of station data and is available from 1880 to 2019.
        This dataset is considered an experimental product, so we use it only for comparative purposes.
  \item To complement blended gridded data products, we use station temperature data from the Global Historical Climatology Network (GHCN) dataset compiled by the National Ocean and Atmospheric Administration \cite{Menne:2012hk} and available at \url{https://www1.ncdc.noaa.gov/pub/data/ghcn/daily/}.
        This dataset provides daily mean, maximum, and minimum temperature observations.
        These measurements represent point measurements, which can differ in important ways from gridded products describing spatial averages due to the spatial heterogeneity of temperature fields.
        We retain stations within the state of Texas if they provide at least 60 years of data and if they contain observations for the set of historical cold extremes shown in \cref{fig:historic_era5}.
\end{enumerate}
We also use population density data from the GPWv4 dataset \cite{ciesin_gpwv4:2016}, a list of power generation facilities from the US Energy Information Administration \cite{useia_generators:2021}, and a map of the Texas Interconnection \cite{useia_regions:2021}.%TODO: add and cite information on the generators that did and did not go out.

\subsection{Inferred per Capita Demand for Heating}\label{sec:inferred-demand}

Most space heating in Texas is either electric or gas \cite{waite_heating:2020} and the majority of power generation in the Texas Interconnection depends on natural gas \cite{everhart_iea:2021}.
Stress on natural gas production and delivery was therefore just as important as the more visible stress on the electric system.

@@

To assess how spatially correlated cold spells might affect the Texas electric grid, we average heating demand in space over the Texas Interconnection domain \cite{useia_regions:2021}, weighting each grid cell by 2020 population density \cite{ciesin_gpwv4:2016}.
We refer to this spatially aggregated time series, which has the straightforward interpretation as the average heating demand experienced by Texas residents, as ``Inferred per Capita Demand for Heating.''

@@

For each event duration considered, return periods are obtained by fitting a stationary generalized extreme value (GEV) distribution to the time series of annual minimum December-February (DJF) temperatures, or to the time series of annual maxima of inferred per capita demand for heating (\cref{sec:inferred-demand}).
Events that occur in December are coded to the following year.
The 2021 winter season was excluded from return period estimates, allowing us to interpret return periods for the February 2021 event as \emph{a priori} estimates.

@@

\section{How extreme was inferred per capita heating demand over the Texas Interconnect?}

The total shock to Texas heating demand is partially determined by the extent to which cold snaps impact multiple population centers simultaneously.
As such, understanding whether there was precedent for a cold snap simultaneously affecting several regions of Texas's grid that today have high population density is critical.
We therefore use our measure of inferred per capita demand for heating (see \cref{sec:inferred-demand}) to represent the aggregate heating demand induced by cold temperatures.
Aggregating historic temperature fields in space using the 2020 population, we answer the question ``what would the aggregate demand for heating have been had historic cold snaps occurred today?''

\Cref{fig:idf_weighted} shows that the intensity, duration, and recurrence intervals of the February 2021 storm are severe but not unprecedented in the historical record.
For example, at the 6 hour duration the December 1989 storm was substantially more intense and other storms including February 1951 were nearly as intense.
At the two day duration, the 2021 and 1989 events were approximately equally intense and other storms including December 1983 were nearly as intense.
The 2011 storm, which caused rolling blackouts and motivated research into the energy system's vulnerability to cold \cite{ferc_outages:2011}, was quite modest by comparison.
The right panel shows statistical return periods for these extreme events.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{ERCOT_HDD_IDF_MLE_popweighted.pdf}
  \caption{
    The inferred per capita demand for heating induced by the February 2021 cold snap is not unprecedented.
    For the worst 6 hours, the 1989 event was more severe than the 2021 event, while they are comparable for the longer durations.
    (a): time series of annual maximum inferred per capita demand for heating (\cref{sec:inferred-demand}) at 6 hour and 2 day durations.
    December extremes, including the December 1989 storm, are coded to the following year so that one maximum per December-February winter season is taken.
    (b): the intensity-duration-frequency intervals estimated using 1950-2020 data (i.e., not using the 2021 event), overlaid by the annual maxima from the 1989, 2011, and 2021 events.
    Gray dashed lines indicate 2, 5, 10, 25, 50, and 100 year return levels.
  }\label{fig:idf_weighted}
\end{figure}

@@

\Cref{fig:local_era5} shows local return periods for the February 2021 cold snap at 6 hour, 1 day, 2 day, and 4 day durations.
Other than a band from south-central to south-east Texas, nearly all regions of the Texas Interconnection (gray outline in \cref{fig:local_era5}a,d) experienced cold with a return period below 50 years.
Results are similar using station data (\cref{fig:local_ghcnd}).
Importantly for the energy system, the band experiencing cold with return period greater than 50 years includes a substantial fraction of Texas’s population (\cref{fig:local_era5}a) and natural gas generation (\cref{fig:local_era5}d).
Outside the Texas Interconnection region, much of the Texas Panhandle and central Oklahoma experienced intense cold with a return period greater than 100 years at multiple durations.
Although Oklahoma recorded some water main breaks \cite{crum_water:2021} and rolling blackouts for several hours \cite{money_oklahoma:2021}, infrastructure impacts were less severe than in Texas despite recording more extreme (relative both to its own historical record and in absolute terms) temperatures.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{local_rt_era5.pdf}
  \caption{
    Return periods for the February 2021 event, calculated using stationary estimates of annual extremes over the period 1950-2020.
    Return periods are calculated separately for each cell.
    (a): estimates of 2020 population density \cite{ciesin_gpwv4:2016}.
    (d): energy generation facilities in Texas \cite{useia_generators:2021}.
    (b,c,e,f): local return periods for 6 hour, 1 day, 2 day, and 4 day durations, respectively.
    Contours enclose regions that recorded 50 and 100 year return levels.
    The gray region in panels (a) and (d) shows boundaries of the Texas Interconnection \cite{useia_regions:2021}.
  }\label{fig:local_era5}
\end{figure}

@@

Our spatially aggregated metric of inferred per capita demand for heating shows that the February 2021 event was intense but not unforeseeable (\cref{fig:idf_weighted}).
Although specific locations experienced very intense ($>100$ year return period) temperatures, we find that for most locations in Texas the temperatures recorded during the February 2021 cold snap had precedent in the historical record.

A proximate cause of load shedding in the Texas Interconnection during February 2021 was the vulnerability of the electricity generation system to the severe cold temperatures \cite{everhart_iea:2021}.
\Cref{fig:local_era5,fig:local_ghcnd} show that this happened even though most parts of the state had previously experienced similarly intense cold, notably in 1989.
While our analysis neglects other meteorological factors, like freezing rain, that may have impeded operations at specific facilities, our analysis suggests that the February 2021 failures of energy and electricity systems in the Texas Interconnection took place during temperatures comparable to those previously recorded.
Similarly, water mains broke in places like Houston where temperatures did not exceed 100-year return levels, underscoring characteristic vulnerability of critical infrastructure systems \cite{chester_reliable:2020}.

@@

Our analysis quantifies the frequency with which the temperatures observed during February 2021 could have been expected to occur \emph{a priori}.
Other factors also govern infrastructure performance and failure, including precipitation, the demand for natural gas in adjacent regions, and complex connections within and between regional systems.
Similarly, decisions at multiple time scales, including disaster preparedness and risk communication, contribute to the human consequences of physical infrastructure failure.
Thus, the exact chain of events that led to the blackouts and water system disruptions during February 2021 should be sorted out only after thorough investigations by parties on the ground in Texas.

@@

The February 2021 cold snap was the most intense in 30 years, but was not without precedent in the full historical record.
In addition to the record cold conditions of 1899 (\cref{fig:historic_bk}), we estimate that the weather of December 1989 would have resulted in slightly higher 6-hour and 2-day aggregate heating demands over the Texas Interconnection had it occurred in February 2021.
Several other storms since 1950 would have produced nearly as much demand for heating.
Given upward trends in the electrification of heating, it is likely that future cold snaps will cause peak annual loads on the Texas Interconnection to occur during the winter season.
Infrastructure expansion necessitated by a rapidly growing population offers Texas the opportunity to invest in a more resilient energy system.

@@

\subsection{Inferred per capita demand for heating}

To complement our analysis of inferred per capita demand for heating, we consider how results change as a function of two modeling decisions.
First, we consider what happens if the spatial field demand for heating is aggregated using grid cell area rather than population density.
Next, we compute return periods using an estimator based on the method of $L$-moments.
Although $L$-moment estimators for the generalized extreme value distribution are not unbiased, they are popular in the statistical hydrology literature for their stability \cite{hosking_gev:1985,martins_gev:2001,morrison_gev:2002}.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{HDD_pop_weighted_ts.pdf}
  \caption{
    A time series of inferred per capita demand for heating over the Texas Interconnection during the February 2021 cold snap.
  }\label{fig:hdd_ts}
\end{figure}
